---
title: "ESM204 HW3"
author: "Taylor Lockmann & Amber McEldowney"
date: "5/9/2021"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

options(scipen=999) # not scientific notation
gc() # clear RAM

library(tidyverse)
library(stringr)
library(stargazer)
library(cowplot)
library(here)
library(patchwork)
```

## 1. Marginal externality cost (MEC) per gallon

```{r}
mec_gallon <- 51*(1/2000)*19.6
returnValue(mec_gallon)
```
Marginal externality cost is approximately $0.50 per gallon of gas. 


## 2. Aggregate demand curve, supply curve, consumer benefit, producer benefit, and environmental cost under status quo.

```{r}
# Read in the data

demand_data <- read.csv(here("data", "HW3_data.csv"), stringsAsFactors = F)

# Exploratory plots
plot_1 <- ggplot(data=demand_data, aes(x=Q_Low_Gallons,y=Price_Dollars))+
  geom_point()+
  theme_cowplot(16)

plot_2 <- ggplot(data=demand_data, aes(x=Q_High_Gallons,y=Price_Dollars))+
  geom_point()+
  theme_cowplot(16)

#plot_1 + plot_2
```

```{r}
# Estimate linear models

## Low income group linear models:

low_demand_model <- lm(Price_Dollars ~ Q_Low_Gallons, data = demand_data)
# low_demand_model$coefficients # Call the coefficients

al <- low_demand_model$coefficients[[1]] # "al" is the "a" coeff. for low demand
bl <- low_demand_model$coefficients[[2]] # "bl" is the "b" coeff. for low demand

## High income group linear models:

high_demand_model <- lm(Price_Dollars ~ Q_High_Gallons, data = demand_data)
# high_demand_model$coefficients # Call the coefficients

ah <- high_demand_model$coefficients[[1]] # "ah" is the "a" coeff. for high demand
bh <- high_demand_model$coefficients[[2]] # "bh" is the "b" coeff. for high demand
```

```{r}
# Add our estimated curves to our exploratory plots

price_low_fitted <- al + bl*demand_data$Q_Low_Gallons
price_high_fitted <- ah + bh*demand_data$Q_High_Gallons

#(plot_1+
#  geom_line(aes(y=price_low_fitted, x=Q_Low_Gallons))) +
#(plot_2+
#   geom_line(aes(y = price_high_fitted, x = Q_High_Gallons)))
```

```{r}
# Create models to help us find aggregate demand curve

# P(Q)
inverse_demand <- function(q, model){
  p <- model$coefficients[[1]] + model$coefficients[[2]]*q
return(p)
}

## Inverse P(Q) to get Q(P)
demand <- function(p, model){
  q <- (p - model$coefficients[[1]])/model$coefficients[[2]]
  return(q)
}

# Make aggregate demand model
agg_demand_model <- lm(Price_Dollars ~ Q_High_Gallons+Q_Low_Gallons, data = demand_data)
# agg_demand_model$coefficients # Call the coefficients

at <- agg_demand_model$coefficients[[1]] # "at" is the "a" coeff. for total demand
bt <- agg_demand_model$coefficients[[2]] # "bt" is the "b" coeff. for total demand

# Find q given price p ($3)
Q_star <- demand(3, agg_demand_model)

# Make supply model
supply <- function(p, agg_demand_model){
  q <- (p - agg_demand_model$coefficients[[1]])/agg_demand_model$coefficients[[2]]
 return(q)
}
```


```{r}
# Now create functions for our lines

low_demand_curve <- function(x) low_demand_model$coefficients[[1]] + low_demand_model$coefficients[[2]]*x
high_demand_curve <- function(x) high_demand_model$coefficients[[1]] + high_demand_model$coefficients[[2]]*x
agg_demand_curve <- function(x) agg_demand_model$coefficients[[1]] + agg_demand_model$coefficients[[2]]*x

# Find slope of supply curve (we know y-intercept is 0, and current price is $3)
bs <- 3/supply(3, agg_demand_model) ## rise over run
supply_curve <- function(x) bs*x

# Make a MSC curve from supply curve + MEC curve (MEC found in part 1)
msc_curve <- function(x) bs*x + 0.4998


# Make the plot
ggplot(data = demand_data, aes(y = Price_Dollars, x = Q_High_Gallons))+
  stat_function(fun = low_demand_curve,
                color = "red")+
  stat_function(fun = high_demand_curve,
                color = "green")+
  stat_function(fun = supply_curve,
                color = "blue")+
  stat_function(fun = msc_curve,
                color = "pink")+
  annotate("text", x = 400000, y = 14, color = "red", label = "Low Income Group")+
  annotate("text", x = 400000, y = 16, color = "green", label = "High Income Group")+
  annotate("text", x = 400000, y = 12, color = "orange", label = "Aggregate Demand")+
  annotate("text", x = 770000, y = 3, color = "blue", label = "Supply Curve")+
  annotate("text", x = 700000, y = 5, color = "pink", label = "MSC")+
  annotate("text", x = 800000, y = 1, color = "purple", label = "MEC")+
  annotate("text", x = -5, y = 3, color = "black", label = "P*")+
  annotate("text", x = Q_star, y = -1, color = "black", label = "Q*")+
  theme_bw()+
  labs(
    y = "Price (USD)",
    x = "Quantity (gallons)",
    title = "Demand Curves for Gasoline"
  )+
  geom_hline(aes(yintercept = 0))+
  geom_vline(aes(xintercept = 0))+
  theme(plot.title = element_text(hjust = 0.5))+
  ylim(-1, 20)+
  xlim(-5, 800000)+
  geom_segment(x = 0, y = 15.803, xend = 150807, yend = 11.685, color = "orange")+ # Use above models to find the y-intercepts of the two demand curves. This will be the y and yend of our pre-kink line segment. 
  geom_segment(x = 150807, y = 11.685, xend = 801395.3, yend = 0, color = "orange")+ # Use models above to find the x-intercept of the aggregate demand curve. 
    geom_segment(linetype = "dashed", col = "black", x = 0, y = 3, xend = Q_star, yend = 3)+
  geom_segment(linetype = "dashed", col = "black", x = Q_star, y = 0, xend = Q_star, yend = 3)+
  geom_hline(aes(yintercept = 0.4998), color = "purple")
```



The benefit to consumers under the status quo will be equal to the consumer surplus. We can approximate this as follows: 

```{r}
CS_p <- function(p, model){
  q <- demand(p, model)
  cs <- 0.5*(model$coefficients[[1]] - p)*q
  return(cs)
}

CS_p(3, agg_demand_model)
```

Therefore we see that the consumer surplus is approximately $3,741,094. 
**Note:** It is worth noting that this is not an exact value, as this is the area under the aggregate demand curve from 0 to Q*. In the above graph we can see that the aggregate demand curve is actually kinked, and is therefore not a perfect triangle, so our obtained value is actually a slight overestimation.


The benefit to producers will be equal to producer surplus. This is as follows:

```{r}
PS_p <- function(p, model){
  q <- supply(p, model)
  ps <- 0.5*p*q
  return(ps)
}

PS_p(3, agg_demand_model)
```

The producer surplus is approximately $957,188.60. We can see that this is much less than the consumer surplus.


The environmental cost under the status quo is as follows:

```{r}
env_cost <- mec_gallon*Q_star
returnValue(env_cost)
```

The environmental cost is equal to $318,935.30.


## 3. Consumer benefit between high and low income consumers

Just looking at the graph in part 2, we can see that the lower income group holds a smaller share of consumer benefit than the higher income group.

```{r}
CS_p(3, high_demand_model)
CS_p(3, low_demand_model)
```

These computed values show us that the lower income group has approximately $ `r CS_p(3, high_demand_model) - CS_p(3, low_demand_model)` less consumer surplus than the higher income group.


## 4. Gas Tax

Since our MEC is equal to $`r mec_gallons`, this would be our optimal gas tax. This is the amount of tax that should be added to every gallon to make our Supply curve equal to our Marginal Social Cost (MSC) curve. 

